---
title: "Lab 17"
author: "Marriane Alahwerdi (A16902759)"
format: html
---

## Notes, HW BELOW 

tar -xvf: to untar files 
gunzip: un zip files
cd: change dir 
 
The PATH is a special environment variable that stores the location of exe files in UNIX systems. I can print it with "echo $PATH" 
 we can add it with: 
 export PATH=$PATH:/my/new/path 
 
 The SRA-TOOLKIT allows us to query and download raw sequencing data from NCBI's SRA database. Key commands include 
 
 prefetch [SRACODE]
 fasta-dump [SRACODE]
 
 ssh -i "~/Downloads/bimm143_ma.pem" ubuntu@ec2-54-202-82-208.us-west-2.compute.amazonaws.com
 
ls *_quant
 
scp -r -i ~/Downloads/bimm143_ma.pem ubuntu@ec2-18-237-254-37.us-west-2.compute.amazonaws.com:~/*_quant . 
 
 
## Downstream analysis

```{r}
getwd()
```

```{r}
print(folders)
```

```{r}
print(files)
```
```{r}
file.exists(files)
```


```{r}
library(tximport)

# Setup the folder and filenames to read
folders <- dir(pattern = "SRR21568*")
if (length(folders) == 0) stop("No folders found with pattern 'SRR21568*'.")

samples <- sub("_quant", "", folders)
files <- file.path(folders, "abundance.h5")
names(files) <- samples

# Import with tximport
```

```{r}
head(txi.kallisto$counts)
```


 
