---
title: "Lab 10: Halloween Mini-Project"
author: "Marriane Allahwerdi (A16902759)"
format: pdf
---

```{r}
candy_file <- "candy-data.csv"
candy <- read.csv(candy_file, row.names = 1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

There are 85 different types of candy in the dataset.


> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

There are 38 fruity candies in the dataset. 

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

81.64291 

```{r}
candy["Twix",]$winpercent
```


> Q4. What is the winpercent value for “Kit Kat”?

76.7686

```{r}
candy["Kit Kat",]$winpercent
```


> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

49.6535

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```


```{r}
library(dplyr)
candy |>
  filter(rownames(candy)=="Haribo Happy Cola") |>
  select(winpercent)
```


```{r}
candy |>
  filter(winpercent > 50) |>
  filter(fruity==1)
```

```{r}
candy[candy$winpercent > 50,][candy$fruity==1,]
```


To get a quick insight into a new dataset some folks like using the skimer package and its `skim()` function.

```{r}
library("skimr")
skim(candy)
```

```{r}
skimr::skim(candy)
```



> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Winpercent column has a range that is much larger than between 0-1.It is measured on a different scale than everything else! I will need to scale my data before doing analysis. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

There is no chocolate in that column. 

>Q8. Plot a histogram of winpercent values

We can do this a few ways, e.g. the "base" in R `hist()` function or `ggplot()`

```{r}
hist(candy$winpercent, breaks=30)
```
```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(binwidth=8) +
  theme_bw()
```

>Q9. Is the distribution of winpercent values symmetrical?

No, the distribution is not very symmetrical.

>Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

The median is below 50% (47.8%) so the center of distribution is below 50%. 

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Chocolate candy is higher ranked than fruit candy. 

```{r}
fruit.candy <- candy |>
  filter(fruity==1) 

summary(fruit.candy$winpercent)
```

```{r}
choc.candy <- candy |>
  filter(chocolate==1) 

summary(choc.candy$winpercent)
```

```{r}
summary(candy[as.logical(candy$chocolate),]$winpercent)
```


>Q12. Is this difference statistically significant?

```{r}
t.test(choc.candy$winpercent, fruit.candy$winpercent)
```

P-Value < 0.05 (p-value = 2.871e-08) so it is statistically signifigant. 

> Q13. What are the five least liked candy types in this set?

The 5 least liked candies are: Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.  

```{r}
play <- c("d", "a", "c")
sort(play)
order(play)
play[order(play)]
```

```{r}
head(candy[order(candy$winpercent), ], 5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

The top 5 candies are Snickers, Kit Kat, Twix, Reese's Miniatures, and Reese's Peanut Butter cups. 

```{r}
tail(candy[order(candy$winpercent), ], 5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

```{r}
ggplot(candy) + 
  aes(x=winpercent, y=reorder(rownames(candy),winpercent),
  fill=chocolate) +
  geom_col()
```

I want a more specialized, costom color scheme where I can see both chocolate and bar and fruit all from one plot. To do this we can roll our own color vector.

```{r}
mycols <- rep("black", nrow(candy))
mycols[as.logical(candy$chocolate)] <- "chocolate"
mycols[as.logical(candy$bar)] <- "brown"
mycols[as.logical(candy$fruity)] = "pink"
mycols
```

```{r}
mycols[rownames(candy)=="Twix"] <- "blue"
```


```{r}
ggplot(candy) + 
  aes(x=winpercent, y=reorder(rownames(candy),winpercent),
  fill=chocolate) +
  geom_col(fill=mycols)

```

```{r}
mycols <- rep("black", nrow(candy))
mycols[rownames(candy)=="Twix"] <- "blue"

mycols[as.logical(candy$chocolate)] <- "chocolate"
mycols[as.logical(candy$bar)] <- "brown"
mycols[as.logical(candy$fruity)] = "pink"
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets. 

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst. 

```{r}
mycols[as.logical(candy$fruity)] <- "red"
```


```{r}
ggplot(candy) +
  aes(winpercent, pricepercent) + 
  geom_point(col=mycols)
```
Add labels

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) + 
  geom_point(col=mycols) +
  geom_label()
```

Make the labels non-overlapping 

```{r}
library(ggrepel)
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) + 
  geom_text_repel(col=mycols, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reece's miniatures 

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

The top 5 most expensive candies in the dataset are: Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate. The least popular is Nip L Nip. 



```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
cij
```

```{r}
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruit and chocolate are anti-correlated. 

> Q23. Similarly, what two variables are most positively correlated?

The two most positively correlated are chocolate and winpercent.


## PCA: Principle Component Analysis

```{r}
pca <- prcomp(candy, scale=T)
summary(pca)
```

```{r}
PCA <- plot(pca$x[,1:2], col=mycols, pch=16)
```

How do original columns contribute to the new PCs. I will look at PC1 here.

```{r}
pca$rotation
```
```{r}
loadings <-as.data.frame(pca$rotation)
```


```{r}
loadings <-as.data.frame(pca$rotation)
ggplot(loadings) + 
  aes(PC1, reorder(rownames(loadings), PC1), fill=PC1) +
  geom_col()
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```


```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=mycols)

p
```


```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```


> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, pluribus, and hard are picked up strongly by PC1 and this makes sense because these most popular kinds of candies are fruity and chocolate candies in boxes and bags. 
