---
title: "Lab 12 BIMM 143"
author: "Marriane Allahwerdi (A16902759)"
format: pdf
---

## Section 4: Population Scale Analysis

You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 expression.
The first column is sample name, the second
column is genotype and the third column are the expression values.

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("4-up on 10-29-24 at 10.04 AM #9 (compiled).txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```


> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?


```{r}
library(ggplot2)
ggplot(expr) + aes(x=geno, y=exp, fill=geno) + geom_boxplot(notch=T)
```






